---
title: "Panel Data & Fixed Effects models"
author: "Lucas Owen"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(questionr)
library(stargazer)
library(tidyverse)
library(estimatr)
rm(list = ls())
```



# Prerequisite

```{r}

# load data
gapminder <- read.csv("data/gapminder2.csv", na.strings=c(-1))
polity <- read_csv("data/polity2.csv")
```


# Explore missing data

Missing data may introduce selection bias. Thus, during cleaning, analyze the source of missingness. Merge datasets `gapminder2.csv` and `polity2.csv`, then report the amount of missing data.


```{r}

# merge data, but first check out the names¡

# change the names¡

# now merge

merged_df <- left_join(gapminder, polity,
                   by = c("cntry"="country","year"="year"))

# create two new variables:

## one with the log of gdpPercap
## one dividing gdpPercap by 1000

merged_df$ln_gdpPercap <- log(merged_df$gdpPercap)
merged_df$gdpPercap <- merged_df$gdpPercap/1000


# check missigness
questionr::freq.na(merged_df)

# For a single variable
mean(is.na(merged_df$lifeExp))

# check missigness for each variable
df_na <- ifelse(is.na(merged_df), 1, 0)
colMeans(df_na)

merged_df$lifeExp[is.na(merged_df$lifeExp)] <- NA_character_


temp <- na.omit(merged_df) #here's why you don't do this! what if we only care about the relationship between polity & gdp?
nrow(merged_df)
nrow(temp)

temp <- merged_df[,!colnames(merged_df) %in% c("lifeExp")]

```

#let's take a look at our data

```{r}
#remove stuff we don't need
rm(df_na, gapminder, polity, temp)
gc() #garbage can to clear RAM

ggplot(data=merged_df, aes(x=gdpPercap, y=lifeExp))+
  geom_point()+
  theme_bw() #let's add a theme!

ggplot(data=merged_df, aes(x=pop, y=gdpPercap))+
  geom_point()+
  theme_bw() #let's add a theme!

length(unique(merged_df$cntry))
```

#lots of data, but only 142 countries...

```{r}

sample_countries <- c("Afghanistan", "Vietnam", "United States", "Germany", "Tanzania")
ggplot(data=merged_df[merged_df$cntry %in% sample_countries,], aes(x=year, y=ln_gdpPercap))+
  geom_line()+
  facet_wrap(~cntry)+
  theme_bw() #let's add a theme!
```

#AUTOCORRELATION!


# Statistical analysis


## Pooled model


In this section, we will fit a pooled model, formally:

$$
Y_{it} = \alpha + \beta X_{it} + \epsilon_{it} 
$$

Run this pooled model in three different specifications, each with `lifeExp` as the outcome variable. In the first model, regress the outcome on `gdpPercap`; in the second, regress it on `ln_gdpPercap`; and in the third, regress it on both `ln_gdpPercap` and `polity`.

```{r}
lm1 <- lm(lifeExp ~ gdpPercap, data=merged_df)
lm2 <- lm(lifeExp ~ ln_gdpPercap, data=merged_df)
lm3 <- lm(lifeExp ~ ln_gdpPercap + polity, data=merged_df)

```


```{r,results='asis', warning=F}
stargazer(lm1, lm2, lm3,
          type="text")
```



## Because of autocorrelation, our standard errors are going to be overly small! We need to correct them

```{r, warning=F}

# Run robust regressions
lm4 <- lm_robust(lifeExp ~ gdpPercap, data=merged_df, clusters=cntry, se_type="stata")
lm5 <- lm_robust(lifeExp ~ ln_gdpPercap, data=merged_df, clusters=cntry, se_type="stata")
lm6 <- lm_robust(lifeExp ~ ln_gdpPercap + polity, data=merged_df, clusters=cntry, se_type="stata")

stargazer(lm4, lm5, lm6) #doesn't work!

#let's compare
summary(lm1)
summary(lm4)
#only thing that's different is the standard errors

# Create standard lm objects for stargazer compatibility
lm4_standard <- lm(lifeExp ~ gdpPercap, data=merged_df)
lm5_standard <- lm(lifeExp ~ ln_gdpPercap, data=merged_df)
lm6_standard <- lm(lifeExp ~ ln_gdpPercap + polity, data=merged_df)

# Use stargazer with overridden coef and se
stargazer(lm4_standard, lm5_standard, lm6_standard, type="text") #here's the starting point... we're just going to update the standard errors
stargazer(lm4_standard, lm5_standard, lm6_standard,
          se = list(c(lm4$std.error[2]), 
                    c(lm5$std.error[2]), 
                    c(lm6$std.error[2], lm6$std.error[3])),
          type = "text")


```

## Country fixed effects model

Now, we will employ a **country fixed effects** model. Recall that our units ($N$) are countries, and for each country we have several observations that refer to a particular period of time ($T$). Formally:

$$
Y_{it} = \alpha_i + \beta X_{it} + \epsilon_{it} 
$$


By including the variable `country`, the model will automatically generate dummy variables for each country within the variable. The coefficients associated with each country represent the country-specific intercepts. While there are situations where these coefficients can be interpreted, it is not always necessary to do so.

```{r, warning=F}
# Run robust regressions
lm4 <- lm_robust(lifeExp ~ gdpPercap + cntry, data=merged_df, clusters=cntry, se_type="stata")
lm5 <- lm_robust(lifeExp ~ ln_gdpPercap + cntry, data=merged_df, clusters=cntry, se_type="stata")
lm6 <- lm_robust(lifeExp ~ ln_gdpPercap + polity + cntry, data=merged_df, clusters=cntry, se_type="stata")

# Create standard lm objects for stargazer compatibility
lm4_standard <- lm(lifeExp ~ gdpPercap + cntry, data=merged_df)
lm5_standard <- lm(lifeExp ~ ln_gdpPercap + cntry, data=merged_df)
lm6_standard <- lm(lifeExp ~ ln_gdpPercap + polity + cntry, data=merged_df)

# Use stargazer with overridden coef and se
stargazer(lm4_standard, lm5_standard, lm6_standard,
          se = list(c(lm4$std.error[2]), 
                    c(lm5$std.error[2]), 
                    c(lm6$std.error[2], lm6$std.error[3])),
          omit = "cntry",
          add.lines = list(c("Country Fixed Effects", "Yes", "Yes", "Yes")), 
          type = "text")
```


## 2-way fixed effects model

If in addition to country-specific intercepts ($\alpha_i$) we can also include period-specific dummies ($\gamma_{t}$), then we are running a **Two-Way Fixed Effects model**. Formally:


$$
Y_{it} = \alpha_{i}  +\gamma_{t} + \beta X_{it} + \epsilon_{it} 
$$


**IMPORTANT**: Ensure that you treat the period variable, year, as a categorical factor. Otherwise, the variable will not be split into dummy variables for each specific year.


```{r, warning=F}
# turn year into factor:

merged_df$year <- as.factor(merged_df$year)

# Run robust regressions
lm4 <- lm_robust(lifeExp ~ gdpPercap + cntry + year, data=merged_df, clusters=cntry, se_type="stata")
lm5 <- lm_robust(lifeExp ~ ln_gdpPercap + cntry + year, data=merged_df, clusters=cntry, se_type="stata")
lm6 <- lm_robust(lifeExp ~ ln_gdpPercap + polity + cntry + year, data=merged_df, clusters=cntry, se_type="stata")

# Create standard lm objects for stargazer compatibility
lm4_standard <- lm(lifeExp ~ gdpPercap + cntry + year, data=merged_df)
lm5_standard <- lm(lifeExp ~ ln_gdpPercap + cntry + year, data=merged_df)
lm6_standard <- lm(lifeExp ~ ln_gdpPercap + polity + cntry + year, data=merged_df)

# Use stargazer with overridden coef and se
stargazer(lm4_standard, lm5_standard, lm6_standard,
          se = list(c(lm4$std.error[2]), 
                    c(lm5$std.error[2]), 
                    c(lm6$std.error[2], lm6$std.error[3])),
          omit = c("cntry", "year"),
          add.lines = list(c("Country Fixed Effects", "Yes", "Yes", "Yes"), c("Year Fixed Effects", "Yes", "Yes", "Yes")), 
          type = "text")
```

